---
title: "EventProjection: Technical Tutorial and User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EventProjection: Technical Tutorial and User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## 1. Introduction

`EventProjection` is a unified package for enrollment and event projection in clinical trials. It supports:

- **design-stage projection**, where enrollment, event, and dropout models are specified through prior distributions;
- **analysis-stage projection before enrollment completion**, where observed enrollment and time-to-event data are combined with optional prior information;
- **analysis-stage projection after enrollment completion**, where only future events from ongoing participants need to be projected;
- **standard non-cure models** inherited from the `eventPred` workflow;
- **mixture-cure and delayed-treatment-effect extensions** adapted from `EventPredInCure`.

The merged package is designed so that:

- non-cure, non-delay workflows remain as close as possible to `eventPred`;
- cure-aware and delayed-treatment-effect workflows remain as close as possible to `EventPredInCure`;
- users can work through one consistent API centered on `summarizeObserved()`, `fitEnrollment()`, `fitEvent()`, `fitDropout()`, `predictEnrollment()`, `predictEvent()`, and `getPrediction()`.

This vignette is intentionally more technical than a quick-start guide. It is written to serve as the main package tutorial and to document the modeling assumptions, parameterizations, algorithms, interfaces, and current implementation limitations.

```{r}
library(EventProjection)
```

## 2. Package workflow in one page

A typical analysis-stage workflow is:

1. Inspect the data with `summarizeObserved()`.
2. Fit one or more enrollment models using `fitEnrollment()` if enrollment is ongoing.
3. Fit one or more event models with `fitEvent()`.
4. Fit one or more dropout models with `fitDropout()`.
5. Compare candidate models using plots and information criteria.
6. Project enrollment with `predictEnrollment()`.
7. Project events with `predictEvent()`.
8. For a one-step workflow, use `getPrediction()`.

A typical design-stage workflow is:

1. Build prior objects for enrollment, event, and optionally dropout.
2. Pass them to `predictEnrollment()`, `predictEvent()`, or `getPrediction()`.
3. Use simulation outputs and prediction intervals to summarize timing uncertainty.

## 3. Input data structure

For analysis-stage workflows, the subject-level data frame typically contains the following columns.

### Required for enrollment prediction

- `trialsdt`: trial start date
- `cutoffdt`: data-cut date
- `usubjid`: unique subject identifier
- `randdt`: randomization date or enrollment date

### Additional required columns for event prediction

- `time`: observed event or censoring time from randomization
- `event`: event indicator (`1` = event, `0` = no event by cutoff)
- `dropout`: dropout indicator (`1` = dropout before event, `0` otherwise)

### Optional columns

- `treatment`: treatment coded as `1`, `2`, ...
- `treatment_description`: treatment label
- baseline covariates used in non-cure event or dropout models

The merged implementation standardizes input column names to lower case internally, so users do not need to worry about capitalization differences.

The package includes example data:

```{r}
data(interimData1)
data(interimData2)
data(finalData)
```

Use these datasets to reproduce most examples in this tutorial.

## 4. Statistical framework

### 4.1 Study stages addressed by the package

The package is intended for three common projection settings.

#### Design stage

No observed trial data are required. Enrollment, event, and dropout models are specified through prior distributions or fixed parameter values.

#### Enrollment phase

Enrollment is ongoing. The future event count is driven by both:

- new subjects who have not yet enrolled, and
- ongoing subjects who are already enrolled but event-free at the data cut.

#### Follow-up phase

Enrollment is complete. Projection only needs to account for future events from ongoing subjects.

### 4.2 Event-count decomposition

Let `t0` denote the data-cut time and let `D(t)` denote the cumulative number of observed events by calendar time `t`. A standard decomposition is

\[
D(t) = D(t_0) + Q(t_0, t) + R(t_0, t),
\]

where:

- \(D(t_0)\) is the observed event count at the data cut,
- \(Q(t_0, t)\) is the future event count contributed by **ongoing subjects**,
- \(R(t_0, t)\) is the future event count contributed by **newly enrolled subjects**.

This decomposition is central to `predictEvent()` and `getPrediction()`.

## 5. Enrollment models

The enrollment component models the counting process of subject arrivals over time.

Let \(N(t)\) denote the cumulative number of enrolled subjects by study day \(t\), with mean function \(\mu(t)\) and rate function \(a(t)\). Under a Poisson-process formulation,

\[
N(t_1) - N(t_0) \sim \text{Poisson}\{\mu(t_1)-\mu(t_0)\},
\]

where

\[
\mu(t) = \int_0^t a(u) \, du.
\]

### 5.1 Supported enrollment models

`EventProjection` supports the following enrollment models:

- `"poisson"`
- `"time-decay"`
- `"b-spline"`
- `"piecewise poisson"`

### 5.2 Homogeneous Poisson model

This model assumes a constant enrollment rate:

\[
a(t) \equiv \lambda,
\qquad
\mu(t)=\lambda t.
\]

Parameterization used in the package:

\[
\theta = \log(\lambda).
\]

### 5.3 Time-decay model

The time-decay model assumes a gradually increasing enrollment rate:

\[
a(t) = \frac{\mu}{\delta}\left(1-e^{-\delta t}\right),
\]

with corresponding mean function

\[
\mu(t)=\frac{\mu}{\delta}\left[t-\frac{1}{\delta}\left(1-e^{-\delta t}\right)\right].
\]

Parameterization:

\[
\theta = \{\log(\mu), \log(\delta)\}^\top.
\]

### 5.4 B-spline enrollment model

The B-spline model is used to flexibly capture non-monotone or otherwise irregular enrollment patterns. The daily enrollment rate is modeled as

\[
a(t)=\exp\{B(t)^\top\theta\},
\]

where \(B(t)\) is a B-spline basis with `nknots` inner knots. This model is primarily useful **after the trial has started** and is not intended as a design-stage prior.

### 5.5 Piecewise Poisson model

The piecewise Poisson model assumes a constant enrollment rate within prespecified intervals:

\[
a(t)=\lambda_j, \qquad t \in [u_{j-1},u_j).
\]

Parameterization:

\[
\theta_j = \log(\lambda_j).
\]

The vector `accrualTime` contains the **fixed left endpoints** of the accrual intervals.

### 5.6 Enrollment-time simulation

For future enrollment simulation, the package uses an inverse-transform strategy equivalent to sequentially generating exponential waiting times on the cumulative-rate scale. If \(V_i\) denotes the arrival time of the \(i\)-th future subject, then

\[
V_i = \mu^{-1}\left\{\mu(t_0)+E_1+\cdots+E_i\right\},
\]

where \(E_1,E_2,\ldots\) are independent standard exponential variables.

## 6. Event and dropout models

The event and dropout components are survival models for time from enrollment/randomization to event or dropout.

Let \(T\) denote a generic time-to-event random variable, with survival function

\[
S(t)=P(T>t)
\]

and hazard

\[
h(t)=\lim_{\Delta t\to 0}
\frac{P(t \le T < t+\Delta t \mid T \ge t)}{\Delta t}.
\]

All event models may be used through `fitEvent()` and all standard dropout models through `fitDropout()`. The same high-level ideas apply to dropout, except that cure-aware and delayed-treatment-effect extensions are implemented for the **event** model rather than the dropout model.

### 6.1 Supported non-cure event models

- `"exponential"`
- `"weibull"`
- `"log-logistic"`
- `"log-normal"`
- `"piecewise exponential"`
- `"model averaging"`
- `"spline"`
- `"cox"`

### 6.2 Exponential model

The exponential model assumes constant hazard:

\[
h(t)=\lambda,
\qquad
S(t)=e^{-\lambda t}.
\]

In the non-cure implementation that follows the `eventPred` convention, the exponential model is parameterized on the regression scale as

\[
\log(\lambda)=\theta^\top x.
\]

### 6.3 Weibull model

For the Weibull model,

\[
h(t)=\frac{\kappa}{\lambda}\left(\frac{t}{\lambda}\right)^{\kappa-1},
\qquad
S(t)=\exp\left\{-\left(\frac{t}{\lambda}\right)^\kappa\right\}.
\]

This model accommodates increasing, constant, or decreasing hazard depending on the shape parameter \(\kappa\).

### 6.4 Log-logistic model

The log-logistic model has survival function

\[
S(t)=\frac{1}{1+(t/\lambda)^\kappa},
\]

and can accommodate non-monotone hazard shapes.

### 6.5 Log-normal model

If \(\log(T) \sim N(\mu,\sigma^2)\), then

\[
S(t)=1-\Phi\left(\frac{\log(t)-\mu}{\sigma}\right).
\]

Like the log-logistic model, it can capture non-monotone hazard behavior.

### 6.6 Piecewise exponential model

The piecewise exponential model assumes constant hazard within user-specified intervals. If `piecewiseSurvivalTime = c(u_0, u_1, ..., u_J)`, then

\[
h(t)=\lambda_j, \qquad t\in[u_{j-1},u_j).
\]

This model is especially useful when the analyst has clinically meaningful phases of risk or wants a flexible but interpretable hazard model.

### 6.7 Model averaging

The model-averaging event model combines Weibull and log-normal fits using BIC-based weights:

\[
S(t)=w_W S_W(t)+(1-w_W)S_{LN}(t),
\]

where the Weibull weight is

\[
w_W=
\frac{\exp(-\tfrac{1}{2}BIC_W)}{\exp(-\tfrac{1}{2}BIC_W)+\exp(-\tfrac{1}{2}BIC_{LN})}.
\]

This option is useful when both models are plausible and the analyst wants a more robust parametric extrapolation.

### 6.8 Flexible spline model

The spline model is based on the Royston-Parmar framework. A transformed survival function is modeled as a natural cubic spline of log-time. With \(u=\log(t)\),

\[
g\{S(t)\}=c(u),
\]

where `scale` determines the transformation:

- `scale = "hazard"`: \(g(S)=\log[-\log(S)]\)
- `scale = "odds"`: \(g(S)=\log(1/S-1)\)
- `scale = "normal"`: \(g(S)=-\Phi^{-1}(S)\)

With zero inner knots, the spline model reduces to Weibull, log-logistic, or log-normal depending on the chosen scale.

### 6.9 Cox model

The Cox option fits a proportional-hazards model and then represents the fitted baseline hazard as a piecewise exponential model over observed event times. For projection beyond the last observed event time, the package extrapolates using a weighted average of hazard rates from the final `m` intervals.

This is a practical option when users want semiparametric fitting on the observed time range but still need forward simulation.

### 6.10 Dropout models

`fitDropout()` supports the same standard non-cure model families as `fitEvent()` except that cure-aware and delayed-treatment-effect versions are not included for dropout. In `getPrediction()`, `dropout_model = "none"` is also allowed.

## 7. Parameterization used in the merged package

A major design feature of both upstream packages is the use of parameterizations that improve large-sample normality and make posterior simulation straightforward. The merged package retains that philosophy.

### 7.1 Enrollment parameterization

For non-cure enrollment models, the merged package follows the `eventPred` / `EventPredInCure` conventions:

- Poisson: \(\theta = \log(\lambda)\)
- Time-decay: \(\theta = (\log\mu, \log\delta)^\top\)
- B-spline: spline coefficients
- Piecewise Poisson: \(\theta_j = \log(\lambda_j)\)

### 7.2 Non-cure event/dropout parameterization

For the core non-cure models inherited from `eventPred`, the package uses the same parameterization philosophy documented in the upstream package manual. In particular, the package-level documentation presents non-cure model parameters on a regression/AFT scale such as:

- exponential: \(\log(\lambda)=\theta^\top x\)
- Weibull/log-logistic/log-normal: \(\theta=(\beta^\top,\log\sigma)^\top\)
- piecewise exponential: baseline interval parameters plus optional covariate terms
- model averaging: block-diagonal parameter vector for Weibull and log-normal
- spline: spline basis coefficients and optional covariate effects
- Cox: baseline log-hazard increments plus regression coefficients

### 7.3 Cure-model parameterization

For the Chen-style mixture-cure models adapted from `EventPredInCure`, the merged package uses the cure fraction as the **first** parameter:

\[
\theta_1 = \operatorname{logit}(p_{cure}),
\qquad
p_{cure}=\frac{e^{\theta_1}}{1+e^{\theta_1}}.
\]

The remaining elements depend on the baseline survival family.

#### Exponential with cured population

\[
\theta = \{\operatorname{logit}(p_{cure}), \log(\lambda)\}.
\]

#### Weibull with cured population

\[
\theta = \{\operatorname{logit}(p_{cure}), \log(\kappa), \log(\lambda)\}.
\]

#### Log-logistic with cured population

\[
\theta = \{\operatorname{logit}(p_{cure}), \log(\kappa), \log(\lambda)\}.
\]

#### Log-normal with cured population

\[
\theta = \{\operatorname{logit}(p_{cure}), \mu, \log(\sigma)\}.
\]

#### Piecewise exponential with cured population

\[
\theta = \{\operatorname{logit}(p_{cure}), \log(\lambda_1), \ldots, \log(\lambda_J)\}.
\]

### 7.4 Fixed cure-rate interface

A key merged enhancement is the `cure_rate` argument in `fitEvent()` and `getPrediction()`.

When `cure_rate = NULL`, the cure fraction is estimated from the data or is expected to be supplied explicitly in the prior object.

When `cure_rate` is provided:

- the cure fraction is treated as **fixed**, not estimated;
- the returned `theta` still includes `qlogis(cure_rate)` as the first element for consistency;
- the first row and column of `vtheta` are set to zero, so posterior draws do not perturb the fixed cure fraction;
- when `by_treatment = TRUE`, `cure_rate` may be either a scalar or a vector with one element per arm.

This is the main interface you asked to expose clearly in the vignette.

## 8. Mixture-cure interpretation

For cure-aware event models, the survival function takes the form

\[
S(t)=p_{cure} + (1-p_{cure})S_u(t),
\]

where:

- \(p_{cure}\) is the long-term cured fraction,
- \(S_u(t)\) is the survival function among the uncured subpopulation.

This formulation is appropriate when a clinically meaningful fraction of participants may never experience the event within the scientific time horizon.

## 9. Prediction algorithms

### 9.1 Ongoing subjects

For an ongoing subject enrolled at time \(U_i\), observed at data cut \(t_0\), the underlying event time must satisfy

\[
T_i > t_0-U_i.
\]

The package therefore generates future event times from the **conditional distribution given survival up to the data cut**. Conceptually,

\[
P(T_i > t \mid T_i > t_0-U_i) = \frac{S(t)}{S(t_0-U_i)}.
\]

This conditional simulation is central to event projection after a data cut.

### 9.2 New subjects

For subjects not yet enrolled at the data cut:

1. enrollment times are simulated from the chosen enrollment model;
2. event times are generated from the chosen event model;
3. dropout times are generated from the chosen dropout model, if applicable;
4. event, dropout, and administrative censoring are combined to determine whether the event contributes to the future event count.

### 9.3 Posterior simulation versus fixed-parameter simulation

By default, prediction uses approximate posterior uncertainty:

\[
\theta^{(b)} \sim N(\hat\theta, \hat V_\theta),
\]

or the closest numerically stable equivalent.

If `fix_parameter = TRUE`, the package does **not** perturb the fitted parameters and instead conditions on the point estimate or fixed prior values. This is often useful for deterministic scenario exploration.

## 10. Function-by-function guide

### 10.1 `summarizeObserved()`

Use `summarizeObserved()` as the first stop in any analysis-stage workflow.

```{r}
obs1 <- summarizeObserved(
  df = interimData1,
  to_predict = "enrollment and event",
  showplot = FALSE,
  by_treatment = FALSE
)

obs2 <- summarizeObserved(
  df = interimData2,
  to_predict = "event only",
  showplot = FALSE,
  by_treatment = FALSE
)
```

Typical outputs include:

- trial start and cutoff summaries;
- enrollment duration and current enrollment count;
- current event, dropout, and ongoing-subject counts;
- cumulative enrollment and event data;
- daily enrollment rates;
- Kaplan-Meier plots for time to event and time to dropout.

### 10.2 `fitEnrollment()`

Fit one or more candidate enrollment models and compare them.

```{r}
fit_enroll_bs <- fitEnrollment(
  df = interimData1,
  enroll_model = "b-spline",
  nknots = 1,
  showplot = FALSE
)

fit_enroll_td <- fitEnrollment(
  df = interimData1,
  enroll_model = "time-decay",
  showplot = FALSE
)

fit_enroll_pw <- fitEnrollment(
  df = interimData1,
  enroll_model = "piecewise poisson",
  accrualTime = c(0, 90, 180),
  showplot = FALSE
)
```

### 10.3 `predictEnrollment()`

Use a fitted enrollment model or a prior object to generate future arrival times and summarize the time to target enrollment.

#### Analysis-stage enrollment prediction

```{r}
pred_enroll <- predictEnrollment(
  df = interimData1,
  target_n = 300,
  enroll_fit = fit_enroll_bs$fit,
  lags = 30,
  pilevel = 0.90,
  nreps = 200,
  showsummary = TRUE,
  showplot = FALSE
)
```

#### Design-stage enrollment prediction

```{r}
design_enroll_fit <- list(
  model = "piecewise poisson",
  theta = log(26 / 9 * seq(1, 9) / 30.4375),
  vtheta = diag(9) * 1e-8,
  accrualTime = seq(0, 8) * 30.4375
)

design_enroll_pred <- predictEnrollment(
  target_n = 300,
  enroll_fit = design_enroll_fit,
  pilevel = 0.90,
  nreps = 200,
  showsummary = TRUE,
  showplot = FALSE
)
```

### 10.4 `fitEvent()`

Fit one or more event models to the observed time-to-event data.

```{r}
fit_event_weibull <- fitEvent(
  df = interimData2,
  event_model = "weibull",
  showplot = FALSE
)

fit_event_piecewise <- fitEvent(
  df = interimData2,
  event_model = "piecewise exponential",
  piecewiseSurvivalTime = c(0, 140, 352),
  showplot = FALSE
)

fit_event_ma <- fitEvent(
  df = interimData2,
  event_model = "model averaging",
  showplot = FALSE
)

fit_event_cox <- fitEvent(
  df = interimData2,
  event_model = "cox",
  m = 5,
  showplot = FALSE
)
```

### 10.5 `fitDropout()`

Fit a dropout model analogously.

```{r}
fit_dropout_exp <- fitDropout(
  df = interimData2,
  dropout_model = "exponential",
  showplot = FALSE
)

fit_dropout_piecewise <- fitDropout(
  df = interimData2,
  dropout_model = "piecewise exponential",
  piecewiseDropoutTime = c(0, 180),
  showplot = FALSE
)
```

### 10.6 `predictEvent()`

Use fitted event and dropout models to project the timing of future events.

#### After enrollment completion

```{r}
pred_event_complete <- predictEvent(
  df = interimData2,
  target_d = 200,
  event_fit = fit_event_piecewise$fit,
  dropout_fit = fit_dropout_exp$fit,
  pilevel = 0.90,
  nreps = 200,
  showsummary = TRUE,
  showplot = FALSE
)
```

#### Before enrollment completion

```{r}
fit_event_interim <- fitEvent(
  df = interimData1,
  event_model = "weibull",
  showplot = FALSE
)

fit_dropout_interim <- fitDropout(
  df = interimData1,
  dropout_model = "exponential",
  showplot = FALSE
)

pred_event_interim <- predictEvent(
  df = interimData1,
  target_d = 200,
  newSubjects = pred_enroll$newSubjects,
  event_fit = fit_event_interim$fit,
  dropout_fit = fit_dropout_interim$fit,
  pilevel = 0.90,
  nreps = 200,
  showsummary = TRUE,
  showplot = FALSE
)
```

### 10.7 `getPrediction()`

`getPrediction()` is the main high-level interface. It fits models when needed, generates projections, and returns summarized results.

#### Analysis stage before enrollment completion

```{r}
pred_full_interim <- getPrediction(
  df = interimData1,
  to_predict = "enrollment and event",
  target_n = 300,
  target_d = 200,
  enroll_model = "time-decay",
  event_model = "weibull",
  dropout_model = "exponential",
  pilevel = 0.90,
  nreps = 200,
  showsummary = TRUE,
  showplot = FALSE
)
```

#### Analysis stage after enrollment completion

```{r}
pred_full_complete <- getPrediction(
  df = interimData2,
  to_predict = "event only",
  target_d = 200,
  event_model = "cox",
  m = 5,
  dropout_model = "exponential",
  pilevel = 0.90,
  nreps = 200,
  showsummary = TRUE,
  showplot = FALSE
)
```

#### Design-stage projection from priors

```{r}
event_prior <- list(
  model = "piecewise exponential",
  theta = log(c(0.0533, 0.0309) / 30.4375),
  vtheta = diag(2) * 1e-8,
  piecewiseSurvivalTime = c(0, 6) * 30.4375
)

dropout_prior <- list(
  model = "exponential",
  theta = log((-log(1 - 0.05) / 12) / 30.4375),
  vtheta = 1e-8
)

pred_design <- getPrediction(
  df = NULL,
  to_predict = "enrollment and event",
  target_n = 300,
  target_d = 200,
  enroll_prior = design_enroll_fit,
  event_prior = event_prior,
  dropout_prior = dropout_prior,
  pilevel = 0.90,
  nreps = 200,
  showsummary = TRUE,
  showplot = FALSE
)
```

## 11. Covariate-adjusted non-cure models

Covariates are currently supported for standard non-cure event and dropout models.

To keep the example self-contained, the code below constructs a temporary data set with mock baseline covariates.

```{r}
set.seed(101)
interim_cov <- as.data.frame(interimData2)
interim_cov$age <- round(rnorm(nrow(interim_cov), mean = 62, sd = 8))
interim_cov$sex <- factor(sample(c("Female", "Male"), nrow(interim_cov), replace = TRUE))

pred_cov <- getPrediction(
  df = interim_cov,
  to_predict = "event only",
  target_d = 200,
  event_model = "weibull",
  dropout_model = "exponential",
  covariates_event = c("age", "sex"),
  covariates_dropout = c("age"),
  pilevel = 0.90,
  nreps = 200,
  showsummary = TRUE,
  showplot = FALSE
)
```

## 12. Cure-aware workflows

This section focuses on the cure-model functionality that motivated your update request.

### 12.1 Supported cure-aware event models

The merged package supports the following mixture-cure event models in `fitEvent()` and `getPrediction()`:

- `"exponential with cured population"`
- `"weibull with cured population"`
- `"log-normal with cured population"`
- `"log-logistic with cured population"`
- `"piecewise exponential with cured population"`

In addition, delayed-treatment-effect variants are supported in design-stage prediction workflows:

- `"exponential with cured population and delayed treatment"`
- `"weibull with cured population and delayed treatment"`
- `"log-normal with cured population and delayed treatment"`
- `"log-logistic with cured population and delayed treatment"`
- `"piecewise exponential with cured population and delayed treatment"`

### 12.2 Fitting a cure model with estimated cure rate

```{r}
fit_event_cure_est <- fitEvent(
  df = interimData2,
  event_model = "weibull with cured population",
  showplot = FALSE
)
```

### 12.3 Fitting a cure model with fixed cure rate

This is the new interface that should be documented explicitly.

```{r}
fit_event_cure_fixed <- fitEvent(
  df = interimData2,
  event_model = "weibull with cured population",
  cure_rate = 0.25,
  showplot = FALSE
)
```

In the fitted object:

- `fit_event_cure_fixed$fit$theta[1]` is `qlogis(0.25)`;
- the first row and column of `vtheta` are zero;
- only the uncured-population survival parameters are estimated.

### 12.4 Fixed cure rate by treatment arm

When `by_treatment = TRUE`, a scalar `cure_rate` applies to all arms, while a vector applies arm-specific cure fractions.

```{r}
fit_event_cure_fixed_by_trt <- fitEvent(
  df = interimData2,
  event_model = "weibull with cured population",
  by_treatment = TRUE,
  cure_rate = c(0.20, 0.35),
  showplot = FALSE
)
```

### 12.5 Piecewise exponential cure model with fixed cure rate

```{r}
fit_event_cure_piecewise_fixed <- fitEvent(
  df = interimData2,
  event_model = "piecewise exponential with cured population",
  cure_rate = 0.30,
  piecewiseSurvivalTime = c(0, 120, 240),
  showplot = FALSE
)
```

### 12.6 Event projection with fixed cure rate from observed data

```{r}
pred_event_cure_fixed <- getPrediction(
  df = interimData2,
  to_predict = "event only",
  target_d = 200,
  event_model = "weibull with cured population",
  cure_rate = 0.25,
  dropout_model = "exponential",
  pilevel = 0.90,
  nreps = 200,
  showsummary = TRUE,
  showplot = FALSE
)
```

### 12.7 Design-stage projection with fixed cure rate

When `cure_rate` is supplied, the design-stage `event_prior` for a cure model may omit the cure parameter and only provide the non-cure baseline parameters. The merged code pads the prior internally by inserting `qlogis(cure_rate)` as the first element of `theta` and a zero-variance row/column in `vtheta`.

#### Weibull cure model example

```{r}
event_prior_cure_fixed <- list(
  model = "weibull with cured population",
  theta = c(log(1.2), log(300)),
  vtheta = diag(c(1e-8, 1e-8))
)

pred_design_cure_fixed <- getPrediction(
  df = NULL,
  to_predict = "enrollment and event",
  target_n = 300,
  target_d = 200,
  enroll_prior = design_enroll_fit,
  event_prior = event_prior_cure_fixed,
  dropout_prior = dropout_prior,
  cure_rate = 0.30,
  pilevel = 0.90,
  nreps = 200,
  showsummary = TRUE,
  showplot = FALSE
)
```

#### Piecewise exponential cure model example

```{r}
event_prior_cure_fixed_pw <- list(
  model = "piecewise exponential with cured population",
  theta = log(c(0.0040, 0.0025, 0.0015)),
  vtheta = diag(3) * 1e-8,
  piecewiseSurvivalTime = c(0, 120, 240)
)

pred_design_cure_fixed_pw <- getPrediction(
  df = NULL,
  to_predict = "enrollment and event",
  target_n = 300,
  target_d = 200,
  enroll_prior = design_enroll_fit,
  event_prior = event_prior_cure_fixed_pw,
  dropout_prior = dropout_prior,
  cure_rate = 0.30,
  pilevel = 0.90,
  nreps = 200,
  showsummary = TRUE,
  showplot = FALSE
)
```

### 12.8 Delayed treatment effect at the design stage

The delayed-treatment-effect extensions are intended for **design-stage simulation**, not for fitting delayed-treatment models to observed data in `fitEvent()`.

A typical design-stage specification is arm-specific and passed through `event_prior` when `by_treatment = TRUE`.

```{r}
event_prior_delay <- list()
event_prior_delay[[1]] <- list(
  model = "weibull with cured population and delayed treatment",
  theta = c(qlogis(0.20), log(1.1), log(320), 0, 1),
  vtheta = matrix(0, 5, 5)
)
event_prior_delay[[2]] <- list(
  model = "weibull with cured population and delayed treatment",
  theta = c(qlogis(0.35), log(1.1), log(320), 60, 0.70),
  vtheta = matrix(0, 5, 5)
)
```

The last two elements encode the delayed-effect portion in the EventPredInCure-style design-stage parameterization.

## 13. By-treatment workflows

Set `by_treatment = TRUE` to fit or predict separately by treatment arm.

```{r}
fit_event_by_trt <- fitEvent(
  df = interimData2,
  event_model = "weibull",
  by_treatment = TRUE,
  showplot = FALSE
)

fit_dropout_by_trt <- fitDropout(
  df = interimData2,
  dropout_model = "exponential",
  by_treatment = TRUE,
  showplot = FALSE
)
```

At the design stage, by-treatment priors are typically supplied as one list element per treatment group.

## 14. Model selection and practical guidance

A practical model-selection strategy is:

1. start with `summarizeObserved()`;
2. compare several enrollment/event/dropout models;
3. use fitted plots together with AIC/BIC;
4. favor clinically interpretable models when several fits are similar;
5. use cure models only when a long-term plateau is scientifically plausible;
6. reserve delayed-treatment-effect models for planning scenarios where non-proportional hazards are expected.

Some practical rules of thumb:

- use `piecewise exponential` when hazard phases are clinically meaningful;
- use `model averaging` when Weibull and log-normal are both credible extrapolation candidates;
- use `cox` when semiparametric fitting is desired but forward projection is still required;
- use `fix_parameter = TRUE` for deterministic scenario summaries;
- use `target_t` in `getPrediction()` or `predictEvent()` when you need the predictive probability of reaching the event target by a specific future time.

## 15. Current implementation limitations in the merged package

Users should be aware of the following current constraints.

- Cure-aware event models do **not** currently support `covariates_event`.
- `event_prior_with_covariates` is not supported for cure-aware or delayed-treatment event models.
- Pooling treatment-specific `event_prior` objects into an overall prior is not supported for cure-aware models.
- Delayed-treatment-effect models are intended for design-stage simulation workflows, not observed-data fitting via `fitEvent()`.
- Dropout models do not currently have cure-aware or delayed-treatment-effect extensions.

These limitations are implementation choices in the current merged package rather than general limitations of the underlying statistical literature.

## 16. Reproducibility

For reproducible simulation output:

1. call `set.seed()` immediately before `predictEnrollment()`, `predictEvent()`, or `getPrediction()`; or
2. use `seed.num` where available.

When exact upstream comparison is important, a good working expectation is:

- non-cure, non-delay branches should match `eventPred` most closely;
- cure-aware or delayed-treatment-effect branches should match `EventPredInCure` most closely.

Minor differences may still arise when merged validation, numerical safeguards, or edge-case handling differ.

## 17. Additional internal modeling helpers

The package also includes lower-level helpers that support cure-aware event modeling and simulation, including:

- `Chen_2016_event_time()` and related helpers
- cure-model survival-probability helpers such as `SP_Chen_weibull()`
- cure-model log-likelihood helpers such as `loglik_Chen_weibull()`

These are primarily implementation-level utilities that support the higher-level fitting and prediction interfaces.

## 18. Session information

```{r}
sessionInfo()
```

## 19. References

Bagiella, E. and Heitjan, D. F. (2001). Predicting analysis times in randomized clinical trials. *Statistics in Medicine*, 20, 2055-2063.

Chen, T.-T. (2016). Predicting analysis times in randomized clinical trials with cancer immunotherapy. *BMC Medical Research Methodology*, 16, 1-10.

Royston, P. and Parmar, M. K. B. (2002). Flexible parametric proportional-hazards and proportional-odds models for censored survival data, with application to prognostic modelling and estimation of treatment effects. *Statistics in Medicine*, 21, 2175-2197.

Ying, G.-S. and Heitjan, D. F. (2008). Weibull prediction of event times in clinical trials. *Pharmaceutical Statistics*, 7, 107-120.

Zhang, X. and Long, Q. (2010). Stochastic modeling and prediction for accrual in clinical trials. *Statistics in Medicine*, 29, 649-658.
